---
title: "Validation of the St. Gallen Algorithm for Treatment of Proximal Humerus Fractures"
author: "Angela Moosman, MD, Moritz Tannast, MD, Joseph M. Schwab, MD"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
    toc_float: true
---

# Data Import, Tidying, and Validation

The first thing we need to do is to import our excel spreadsheet and create our data_frame(s).

```{r message=FALSE, warning=FALSE}
library(readxl, quietly = TRUE)
library(dplyr, quietly = TRUE)

SGA_df <- read_excel(
  "Schwab Database SGA.xlsx",
  col_types = c(
    "numeric",
    "date",
    "numeric",
    "text",
    "text",
    "numeric",
    "numeric",
    "date",
    "text",
    "numeric",
    "text",
    "numeric",
    "date",
    "numeric",
    "numeric", 
    "numeric",
    "numeric"
  )
)

SGA_df <- rename(SGA_df,
         event = `unplanned secondary surgery (0-12 months postop)`,
         time = `follow up (months)`,
         group = `following algorithm`)
SGA_df$group <- factor(
    SGA_df$group,
    levels = c(0, 1),
    labels = c("Not following SGA", "Following SGA")
  )

SGA_df

```

Since the original SGA was created based on secondary surgery within 12 months, we will create a second dataframe where any events that occurred after twelve months are converted to non-events, and all time to follow-up are capped at 12 months
```{r message=FALSE, warning=FALSE}
SGA_12mo_df <- SGA_df
SGA_12mo_df$event[SGA_12mo_df$event == 1 & SGA_12mo_df$time > 12] <- 0
SGA_12mo_df$time[SGA_12mo_df$event == 0 & SGA_12mo_df$time > 12] <- 12
SGA_12mo_df
```

# Kaplan-Meier Survival Curves

Now we will generate Kaplan-Meier Survival Curves for the overall database.

```{r message=FALSE, warning=FALSE}

library(survival, quietly = TRUE)
library(ggsurvfit, quietly = TRUE)
library(survminer, quietly = TRUE)
library(Rcpp, quietly = TRUE)

# Overall KM Survival Curve
fit_SGA <- survfit2(Surv(time, event) ~ group, data = SGA_df)
fit_SGA %>% ggsurvfit() +
  labs(title = "Overall Kaplan-Meier Survival Curve for St. Gallen Algorithm",
       y = "Percentage Survival",
       x = "Follow-up time, months") +
  add_censor_mark() +
  add_confidence_interval() +
  theme(legend.position = "bottom") +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent,
    expand = c(0.01, 0)
  ) +
  scale_x_continuous(
    breaks = c(24, 48, 72, 96, 120)
  )
```

Now we can do the same Kaplan-Meier Curve for the data that has been capped at 12 months. We will make some slight changes to the x axis to reflect the shorter timeframe.

```{r message=FALSE, warning=FALSE}
# 12-month KM Survival Curve
fit_SGA_12mo <-  survfit2(Surv(time, event) ~ group, data = SGA_12mo_df)
fit_SGA_12mo %>% ggsurvfit() +
  labs(title = "12 Month Kaplan-Meier Survival Curve for St. Gallen Algorithm",
       y = "Percentage Survival",
       x = "Follow-up time, months") +
  add_censor_mark() +
  add_confidence_interval() +
  theme(legend.position = "bottom") +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent,
    expand = c(0.01, 0)
  ) +
  scale_x_continuous(limits = c(0, 12),
                     breaks = c(3, 6, 9))
```

# Logrank tests for comparison of Survival Curves

Next we can perform Logrank tests on both the overall data, and the 12-month capped data.

```{r message=FALSE, warning=FALSE}
surv_diff <- survdiff(Surv(time, event) ~ group, data = SGA_df)
surv_diff

surv_diff_12mo <- survdiff(Surv(time, event) ~ group, data = SGA_12mo_df)
surv_diff_12mo
```

Finally we can plot the overall survival again with some more information.

```{r}
ggsurvplot(
  fit_SGA,
  pval = TRUE,
  conf.int = TRUE,
  risk.table = TRUE,
  # Add risk table
  risk.table.col = "strata",
  # Change risk table color by groups
  linetype = "strata",
  # Change line type by groups
  surv.median.line = "hv",
  # Specify median survival
  ggtheme = theme_bw(),
  # Change ggplot2 theme
  palette = c("#E7B800", "#2E9FDF")
)
```

And here is that curve plotted for the 12-month capped data

```{r message=FALSE, warning=FALSE}
ggsurvplot(
  fit_SGA_12mo,
  pval = TRUE,
  conf.int = TRUE,
  risk.table = TRUE,
  # Add risk table
  risk.table.col = "strata",
  # Change risk table color by groups
  linetype = "strata",
  # Change line type by groups
  surv.median.line = "hv",
  # Specify median survival
  ggtheme = theme_bw(),
  # Change ggplot2 theme
  palette = c("#E7B800", "#2E9FDF")
)
```
